---
title : Using the NOAA Storm Database to Explore the Impact of Weather Events on 
        US Public Health and Economy
author: "Brian Waismeyer"
date  : "Thursday, April 23, 2015"
output: 
    html_document:
        keep_md  : yes
        theme    : journal
        highlight: zenburn
---

## Synopsis
This is an exploratory analysis of an open database provided by the the U.S. 
National Oceanic and Atmospheric Administration (NOAA). It was guided by and
completed to fulfill requirements of a course on 
[Reproducible Research](https://www.coursera.org/course/repdata) on Coursera 
(April 6, 2015 session).

The database used contains data about storms in the United States along with
key estimates of how these storms impacted the human environment. More 
information about the database can be found 
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

The analysis aimed to address two proposed questions.

1. What types of storm events are most harmful to US population health?
2. What types of events have the most impact on US economy?

**PROVIDE BRIEF SUMMARY WHEN ANALYSIS IS COMPLETE**

## Loading Supporting Resources
```{r}
library(stringr)    # for cleaning EVTYPE character strings
library(dplyr)      # for dataset reformatting
library(ggplot2)    # for making our data visualizations
```

## Data Processing
The dataset for the analysis was provided by Coursera instructors but appeared
to be a faithful copy (or at least subset) of the original NOAA database. For
the purposes of the analysis, no "preprocessing" was assumed except that
conducted during measurement, collection, and database construction.

### Getting the Data
```{r cache = TRUE}
# we quickly verify that we are not overwriting any local files...
if(file.exists("./NOAA_database.csv")) {
    # warn the user if a copy already exists
    stop("Please remove the 'NOAA_database.csv' file from the local working
         directory or - if the file is current - skip this processing step.")
} else {
    # if all is clear, we download the dataset
    download_url <- paste0("http://d396qusza40orc.cloudfront.net/",
                           "repdata%2Fdata%2FStormData.csv.bz2")
    
    download.file(url = download_url,
                  destfile = "./NOAA_database.csv")
    
    # clean up the url from the workspace
    rm(download_url)
}

# read the data into R (note that I prefer not to assume characters = factors)
storms <- read.csv("./NOAA_database.csv", 
                   stringsAsFactors = FALSE)

# clean up the download file
unlink("./NOAA_database.csv")
```
### Reviewing the Data
We take a quick look at the data structure, both to understand it and to insure
it loaded correctly.

```{r}
str(storms)
head(storms)
```

### Tidying the Data
Our main interest is going to be the event type variable (```EVTYPE```) and
how it relates to public health and economic variables. So, we start by making
it a proper factor. This ends up requiring a few steps due to inconsistent
formatting.
```{r}
# first we inspect the unique character strings in the EVTYPE variable
head(unique(storms$EVTYPE))
length(unique(storms$EVTYPE))
```
We discover several cases where there are either leading/trailing whitespace
(e.g., ```"   HIGH SURF ADVISORY"```) or where there are multiple versions of the
same string but with different capitalization (e.g., ```"Thunderstorm"```, 
```"THUNDERSTORM"```).

The NOAA documentation provides no reason to believe differences in 
capitalization or whitespace are meaningful, so we clean these up.

```{r}
# removing leading/trailing whitespace
storms$EVTYPE <- str_trim(storms$EVTYPE)

# make capitalization conistent (all lowercase)
storms$EVTYPE <- tolower(storms$EVTYPE)

# check unique strings
ev_length <- length(unique(storms$EVTYPE))
ev_length
```
The NOAA documentation lists 48 unique event types. We have `r ev_length`.

To figure out what is going on, we can inspect how often specific codes are
actually used in the data.

```{r}
# make a table of how often all event codes were used, sorted most to least
code_freq_table <- sort(table(storms$EVTYPE), decreasing = TRUE)
head(code_freq_table)

# what proportion of codes are used more than once?
more_than_one <- sum(code_freq_table > 1)/length(code_freq_table)
more_than_one

# more than 5 times?
more_than_five <- sum(code_freq_table > 5)/length(code_freq_table)
more_than_five
```

It looks like only a small proportion of the codes were used very often. A 
proper cleaning would require digging through the list and identifying 
mispellings, alternate arrangements, etc. 

For the purposes of this analysis, we will proceed with the current data unless
it becomes apparent that our results are being notably impacted by coding
inconsistency. This saves us work (at least for now) and allows for the 
possibility that there were types of storm events observed that were not
anticipated by NOAA documentation requirements.

## Results
This section will present analyses attemping to address our two key questions.

### What types of storm events are most harmful to US population health?

For this analysis, we restricted "public health" to measurements of immediate
harm to persons. So, for example, fatality and injury data were considered
variables of interest but crop and property damage (which could have longer
term consequences for public health) were not.

We address the question first by simply exploring frequencies of fatalities
(```FATALITIES```) and injuries (```INJURIES```) overall.

```{r}
# we start off with some simple summary statistics: ignoring event type, what
# are typical fatality rates per weather event?
summary(storms$FATALITIES)
sd(storms$FATALITIES)

# injury rates?
summary(storms$INJURIES)
sd(storms$INJURIES)
```
Given that most storms appear to have 0 injuries or fatalities, we take a 
simple approach to grouping our events to get a clear take on how often
fatalities and injuries occur at all

```{r}
# first we add two new factors to our dataset to group our events by number
# of fatalities or injuries
storms$FAT_FACTOR <- sapply(storms$FATALITIES, function(x) 
    if(x == 0) {
      return("None")
    } else if (x > 5) {
        return("Five or More")
    } else {
        return("One to Five")
    }
)

storms$INJ_FACTOR <- sapply(storms$INJURIES, function(x) 
    if(x == 0) {
      return("None")
    } else if (x > 5) {
        return("Five or More")
    } else {
        return("One to Five")
    }
)

# now we turn these into proper, ordered factors
storms$FAT_FACTOR <- factor(storms$FAT_FACTOR, 
                            levels = c("None", "One to Five", "Five or More"),
                            labels = c("None", "One to Five", "Five or More")
                            )

storms$INJ_FACTOR <- factor(storms$INJ_FACTOR, 
                            levels = c("None", "One to Five", "Five or More"),
                            labels = c("None", "One to Five", "Five or More")
                            )
```

### What types of events have the most impact on US economy?

